NumPy---Array Computation for Python

### 2020-01-05 body.tex
Draft data structure section

The \code{ndarray} data structure stores regularly spaced elements of a single
data type in a contiguous block of memory; along with meta-data such as shape,
it allows for the efficient representation of $n$-dimensional arrays.

\fixme{discuss .. vectors / matrices / and then more ...
indexing and slicing, one-liners that would be tens or hundreds of lines of code in
languages such as C.}

More details about the data structure are given in ``The NumPy array:
a structure for efficient numerical computation.''\cite{vanderwalt2011numpy}.

\begin{lstlisting}
>>> import numpy as np
>>> x = np.array([0, 1, 2, 3])
>>> x.dtype
dtype('int64')
>>> x.shape
(4,)

>>> y = 2 * x
>>> y
array([0, 2, 4, 6])

>>> x * y
array([ 0,  2,  8, 18])

>>> x @ y
28
\end{lstlisting}

\begin{lstlisting}
>>> import numpy as np
>>> x = np.array([1, 2])
>>> A = np.array([[0.1, 0.2], [0.3, 0.2]])
>>> A.dtype 
dtype('float64')
>>> A.shape
(2, 2)

>>> A * x
array([[0.1, 0.4],
       [0.3, 0.4]])

>>> A @ x
array([0.5, 0.7])
>>> x @ A
array([0.7, 0.6])
\end{lstlisting}

NumPy provides a large collection of functions for array manipulation
including functions for: creating, reshaping, concatenating, and padding arrays;
searching, sorting and counting data
in arrays; computing elementary statistics, such as the mean, median,
variance, and standard deviation; file I/O; and more.
It also provides extensive support for generating pseudorandom numbers
as well as an assortment of probability distributions.
For historical reasons, NumPy also includes basic functionality for
linear algebra, fast Fourier transforms and windowing,
and polynomial fitting.
We will continue supporting these, but we will not expand beyond them.

>>>>

The \code{ndarray} data structure stores regularly spaced elements of a single
data type in a contiguous block of memory; along with meta-data such as shape,
it allows for the efficient representation of $n$-dimensional arrays.
More details about the \code{ndarray} data structure are given in ``The NumPy array:
a structure for efficient numerical computation''\cite{vanderwalt2011numpy}
and our online documentation.

The need for such a data structure arises often in science.
For instance, they may store measurements from an experiment or simulation
at discrete time points.
Data analysis frequently involves operations $(+, -, \times, /, \log, \arctan, \dots)$ on
the individual measurements.
In programming languages such as C, this would involve looping over every element
to perform these operations.
Array computations generalize these element-wise operations on scalars to
vectors (1-d arrays), matrices (2-d arrays), and higher dimensional arrays (n-d
arrays).
For instance, if \code{x} is an \code{ndarray}, then \code{2 * x} produces
an \code{ndarray} whose values are twice those in \code{x}.
Similarly, if \code{x} and \code{y} have the same shape, then
\code{x * y} produces an \code{ndarray} whose values are the products of
the corresponding entries of \code{x} and \code{y}.
These \emph{vectorized} operations often produce one-liners that would tens
or hundreds of lines of code in languages such as C.

Element-wise operations are not always desired.
Suppose the matrix \code{A} is a 2-d array and the vector \code{x} is 1-d array.
Then \code{A @ x} computes the matrix-vector product $Ax$ and
\code{x @ A} computes the vector-matrix product $x^\top A$.
This matrix multiplication operation extends to higher-dimensions by
treating the array as a stack of matrices.

NumPy also provides a large collection of functions for array manipulation
including functions for: creating, reshaping, concatenating, and padding arrays;
searching, sorting and counting data
in arrays; computing elementary statistics, such as the mean, median,
variance, and standard deviation; file I/O; and more.
It also provides extensive support for generating pseudorandom numbers
as well as an assortment of probability distributions.
For historical reasons, NumPy also includes basic functionality for
linear algebra, fast Fourier transforms and windowing,
and polynomial fitting.
We will continue supporting these, but we will not expand beyond them.
